Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2018-01-15 manuscript/manuscript.Rmd
Use of bookdown

---
title: "Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar"
bibliography: bib/biblio.bib
csl: bib/journal-of-applied-ecology.csl
output: 
  pdf_document:
    citation_package: natbib
    fig_caption: true
    keep_tex: true
---

@@

The island of Madagascar has an unparalleled biodiversity, mainly located in the tropical forests of the island, which is highly
  threatened by anthropogenic deforestation. Scattered forest maps
  from past studies at national level with substantial gaps (due to
  presence of cloud cover on satellite imagery) prevent the analysis
  of long-term deforestation trends in Madagascar. In this study, we
  propose a new approach combining historical (1953-2000) national
  forest-cover maps with recent (2001-2014) global annual tree cover
  loss data to look at six decades (1953-2014) of deforestation and
  forest fragmentation in Madagascar. We produced new forest-cover
  maps at 30~m resolution over the full territory of Madagascar for
  the year 1990, and annually from 2000 to 2014. We estimated that
  Madagascar has lost 44\% of its natural forest cover over the period
  1953-2014 (including 37\% over the period 1973-2014). Natural
  forests cover 8.9 Mha in 2014 (15\% of the national territory) which
  are divided into 4.4 Mha (50\%) of moist forests, 2.6 Mha (29\%) of
  dry forests, 1.7 Mha of spiny forests (19\%) and 177,000 ha (2\%) of
  mangroves. Since 2005, the annual deforestation rate has
  progressively increased in Madagascar to reach 99,000 ha/yr during
  2010-2014 (corresponding to a rate of 1.1\%/yr). This increase is
  probably due to rapid population growth (close to 3\%/yr) and to
  poor law enforcement in the country. Around half of the forest
  (46\%) is now located at less than 100~m from the forest
  edge. Accurate forest-cover change maps can be used to assess the
  effectiveness of past and current conservation programs and
  implement new strategies for the future. In particular, forest maps
  and estimates can be used in the REDD+ framework which aims at
  ``Reducing Emissions from Deforestation and Forest Degradation'' and
  for optimizing the current protected area network.

**Keywords**: biodiversity, climate-change, deforestation, Landsat, Madagascar, tropical forest.

## 1. Introduction

Separated from the African continent and the Indian plate about 165 and 88 million years ago respectively [@Ali2008], the flora and fauna of Madagascar followed its own evolutionary path. Isolation combined with a high number of micro-habitats [@Pearson2009] has led to Madagascar's exceptional biodiversity both in term of number of species and endemism in many taxonomic groups [@Crottini2012; @Goodman2005]. Most of the biodiversity in Madagascar is concentrated in the tropical forests of the island which can be divided into three types: the moist forest in the East, the dry forest in the West and the spiny forest in the South [@Vieilledent2016]. This unparalleled biodiversity is severely threatened by anthropogenic deforestation [@Harper2007; @Vieilledent2013] associated with activities such as slash-and-burn agriculture and pasture [@Scales2011]. Tropical forests in Madagascar also store a large amount of carbon [@Vieilledent2016] and high rates of deforestation in Madagascar are responsible for large CO<sub>2</sub> emissions in the atmosphere [@Achard2014]. Deforestation threatens species survival by directly reducing their available habitat [@Brooks2002; @Tidd2001]. Forest fragmentation can also lead to species extinction by isolating populations from each other and creating forest patches too small to maintain viable populations [@Saunders1991]. Fragmentation also increases forest edge where ecological conditions (such as air temperature, light intensity and air moisture) can be dramatically modified, with consequences on the abundance and distribution of species [@Murcia1995]. Forest fragmentation can also have substantial effects on forest carbon storage capacity, as carbon stocks are much lower at the forest edge than under a closed canopy [@Brinck2017]. Moreover, forest carbon stocks vary spatially due to climate or soil factors [@Saatchi2011; @Vieilledent2016]. As a consequence, accurate and spatially explicit maps of forest-cover and forest-cover change are necessary to monitor biodiversity loss and carbon emissions from deforestation and forest fragmentation, assess the efficiency of present conservation strategies [@Eklund2016], and implement new strategies for the future [@Vieilledent2013; @Vieilledent2016]. Simple time-series of forest-cover estimates, such as those provided by the FAO Forest Resource Assessment report [@Keenan2015] are not sufficient.

Unfortunately, accurate and exhaustive forest-cover maps are not available for Madagascar for the last fifteen years (2000-2015). @Harper2007 produced maps of forest cover and forest cover changes over Madagascar for the years c. 1953, c. 1973, 1990 and 2000. The c. 1953 forest map was derived from the visual interpretation of aerial photography at coarse scale (1/1,000,000). Forest maps for the years c. 1973, 1990 and 2000, were obtained from supervised classification of Landsat satellite images at 60m resolution (for the year 1973) or 30m resolution (for years 1990 and 2000) and can be used to derive accurate estimates (89.5% accuracy reported for the forest / non-forest map of year 2000). Maps provided by @Harper2007 were not exhaustive (due to the presence of clouds in the satellite imagery), e.g. 11 244 km2 are mapped as unknown cover type for the year 2000. Using a similar supervised classification approach as in @Harper2007, more recent maps have been produced for the periods 2000-2005-2010 by national institutions, with the technical support of international environmental NGOs [@MEFT2009; @ONE2013]. Another set of recent forest-cover maps using an advanced statistical tool for classification, the Random Forest classifier [@Grinand2013; @Rakotomalala2015], was produced for the periods 2005-2010-2013 [@ONE2015]. However, these maps are either too old to give recent estimates of deforestation [@MEFT2009; @ONE2013], include large areas of missing information (due to images with more than 20% of clouds) for the 2010 map provided by [@ONE2013], or show large mis-classification in specific areas, especially in the dry and spiny forest domain for which the spectral answer has a strong seasonal behavior due to the deciduousness of such forests (overall accuracy is lower than 0.8 for the dry and spiny forests for the maps produced by @ONE2015). Moreover, the production of such forest maps from a supervised classification approach requires significant resources, especially regarding the image selection step (required to minimize cloud cover) and the training step (visual interpretation of a large number of polygons needed to train the classification algorithm) [@Rakotomalala2015]. Most of this work of image selection and visual interpretation would need to be repeated to produce new forest maps in the future using a similar approach.

Global forest or tree cover products have also been published recently and can be tested at the national scale for Madagascar. [@Kim2014] produced a global forest-cover change map from 1990 to 2000 (derived from Landsat imagery). This product was updated to cover the period 1975-2005 (<http://glcf.umd.edu/data/landsatFCC/>) but forest-cover maps after 2005 were not produced. Moreover, the approach used in @Kim2014 did not accurately map the forests in the dry and spiny ecosystems of Madagascar (see Fig. 8 in @Kim2014). @Hansen2013 mapped tree cover percentage, annual forest loss and forest gain from 2000 to 2012 at global scale at 30 m resolution. This product has since been updated and is now available up to the year 2014 [@Hansen2013]. To map forest cover from the @Hansen2013 product, a tree cover threshold must be selected (that defines forest cover). Selecting such a threshold is not straightforward as the accuracy of the global tree cover map strongly varies between forest types, and is substantially lower for dry forests than for moist forests [@Bastin2017]. Moreover, the @Hansen2013 product does not provide information on land-use. In particular the global tree cover map does not separate tree plantations such as oil palm or eucalyptus plantations from natural forests [@Tropek2014]. Thus, the global tree cover map from @Hansen2013 cannot be used alone to produce a map of forest cover [@Tyukavina2017]. In complement to the tree cover percentage provided in @Hansen2013, a layer of annual tree cover loss is also provided (i.e. complete loss of tree cover from a value higher than 10% to zero) for the period 2001-2014.

In this study, we present a simple approach which combines the maps from @Harper2007 and products from @Hansen2013 to derive annual wall-to-wall forest-cover change maps over the period 2000-2014 for Madagascar. We use the forest-cover map provided by @Harper2007 for the year 2000 (defining the land-use) with the tree cover loss product provided by @Hansen2013 that we apply only inside forest areas identified by @Harper2007. Similar to the approach of @Harper2007, we also assess trends in deforestation rates and forest fragmentation from c. 1953 to 2014. The approach described in this study can help assess the effectiveness of current conservation strategies, and assist the implementation of future strategies. Our approach could be easily extended to other tropical countries that have at least one forest-cover map between 2000 and 2014. This approach can easily be repeated in the future when the @Hansen2013 products are updated.

## 2. Materials and Methods

### 2.1 Creation of new forest-cover maps of Madagascar from 1953 to 2014

We produced annual forest/non-forest maps at 30m resolution for the full territory of Madagascar for the period 2000-2014 by combining the forest map of year 2000 from @Harper2007, and the tree cover percentage and annual forest cover loss maps over the period 2000-2014 from @Hansen2013. The 2000 Harper's forest map includes 208,000 ha of unclassified areas due to the presence of clouds in input satellite imagery, mostly (88%) within the moist forest domain which covers 4.17 million ha in total. To provide a label (forest or non-forest) to these unclassified pixels, we use the 2000 tree cover percentage map of @Hansen2013 by selecting a threshold of 75% tree cover to define forest cover as recommended by other studies for the moist domain [@Achard2014]. We thus obtained a forest-cover map for the year 2000 covering the full territory of Madagascar. We then combined this forest-cover map of the year 2000 with the annual forest cover loss maps from 2001 to 2014 provided by @Hansen2013 to create annual forest-cover maps from 2001 to 2014 at 30m resolution. We also completed the Harper's forest map of year 1990 by filling unclassified areas (due to the presence of clouds on satellite images) using our forest-cover map of year 2000. To do so, we assumed that if forest was present in 2000, the pixel was also forested in 1990. The remaining unclassified pixels were limited to a relatively small total area of c. 8,000 ha. We labeled these residual pixels as non-forest as for the year 2000. Similarly we completed the Harper forest map of year 1973 by filling unclassified areas using our forest-cover map of the year 1990 assuming that if forest was present in 1990, it was also present in 1973. Contrary to the year 1990, the remaining unclassified pixels for year 1973 corresponded to a significant total area of 3.32 million ha. We also reprojected the forest-cover map of year 1953 to a common projection in order to compare to the other eras. This map was produced by scanning a paper map derived from aerial photos, and thus could not be perfectly aligned with the other maps produced through digital processing of satellite imagery [@Harper2007]. Finally for all forest-cover maps from 1973, the isolated single non-forest pixels (i.e. fully surrounded by forest pixels) were removed, assuming that single non-forest pixels inside a forest patch were not corresponding to deforestation (they might correspond to selective logging activities). This allowed us to avoid counting very small scale events (<0.1 ha such as selective logging) as forest fragmentation. All the resulting maps are freely available at <https://bioscenemada.cirad.fr/forestmaps>.

### 2.2 Computing forest-cover areas and deforestation rates

From these new forest-cover maps, we calculated the total forest-cover area for seven available years (1953-1973-1990-2000-2010-2005-2014), and the annual deforested area and annual deforestation rate for the corresponding six time periods between 1953 and 2014. The annual deforestation rates were calculated as follows [@Puyravaud2003; @Vieilledent2013]:

$$\theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}]$$

where $\theta$ is the annual deforestation rate (in %/yr), $F_{t_2}$ and $F_{t_1}$ are the forest cover free of clouds at both dates $t_2$ and $t_1$, and $t_2-t_1$ is the time-interval (in years) between the two dates.

Because of the large unclassified area (3.32 million ha) in 1973, the annual deforestation areas and rates for the two periods 1953-1973 and 1973-1990 are only indicative estimates. For these two periods the annual deforestation rates are computed as the ratio $(F_{t_2}-F_{t_1})/F_{t_1}$ considering only the mapped forest pixels. Area and rate estimates are produced at the national scale and for the four forest ecosystems present in Madagascar: moist forest in the East, dry forest in the West, spiny forest in the South, and mangroves on the Western coast (Fig. 1). To define the forest domains, we used a map from the MEFT (_"Ministère de l'Environnement et des Forêts à Madagascar"_) with the boundaries of the four ecoregions in Madagascar. Ecoregions were defined on the basis of climatic and vegetation criteria using the climate classification by @Cornet1974 and the vegetation classification from the 1996 IEFN national forest inventory [@IEFN1996]. Because mangrove forests are highly dynamic ecosystems that can expand or contract on decadal scales depending on changes in environmental factors [@Armitage2015], a fixed delimitation of the mangrove ecoregion on six decades might not be fully appropriate. As a consequence, our estimates of the forest-cover and deforestation rates for mangroves in Madagascar must be considered with this limitation.

### 2.3 Comparing our forest-cover and deforestation rate estimates with previous studies

We compared our estimates of forest-cover and deforestation rates with estimates from the three existing studies at the national scale for Madagascar: (i) [@Harper2007], (ii) [@MEFT2009] and (iii) [@ONE2015]. @Harper2007 provides forest-cover and deforestation estimates for the periods c. 1953-c. 1973-1990-2000. MEFT, USAID, and CI (2009) provides estimates for the periods 1990-2000-2005 and ONE, DGF, MNP, WCS, and Etc Terra (2015) provides estimates for the periods 2005-2010-2013. To compare our forest-cover and deforestation estimates over the same time periods, we consider an additional time-period in our study (2010-2013) by creating an extra forest-cover map for the year 2013. We computed the Pearson's correlation coefficient and the root mean square error (RMSE) between our forest-cover estimates and forest-cover estimates from previous studies for all the dates and forest types (including also the total forest cover estimates). For previous studies, the computation of annual deforestation rates (in %/yr) is not always detailed and might slightly differ from one study to another [see @Puyravaud2003]. @Harper2007 also provide total deforested areas for the two periods 1973-1990 and 1990-2000. We converted these values into annual deforested area estimates. When annual deforested areas were not reported (for 1953-1973 in @Harper2007 and in @MEFT2009 and @ONE2015), we computed them from the forest-cover estimates in each study. These estimates cannot be corrected from the potential bias due to the presence of residual clouds. Forest-cover and deforestation rates were then compared between all studies for the whole of Madagascar and the four ecoregions. The same ecoregion boundaries as in our study were used in @ONE2015 but this was not the case for @Harper2007 and @MEFT2009, which can explain part of the differences between the estimates.

### 2.3 Fragmentation

We also conducted an analysis of changes in forest fragmentation for the years 1953, 1973, 1990, 2000, 2010 and 2014. We applied the method developed by @Riitters2000 which uses a moving window to characterize the fragmentation around each forested pixel. Computations were done using the function `r.forestfrag` of the GRASS GIS software [@Neteler2008]. Six categories of fragmentation were identified from the amount of forest and its occurrence as adjacent forest pixels: "interior", "perforated", "edge", "transitional", "patch", and "undetermined". We used a moving window of 7x7 pixels (4.4 ha). Using this window size, forest edge had a width of about 90m [@Riitters2000]. The "interior" category can be interpreted as the most intact forest [@Potapov2017]. The "patch" and "transitional" categories correspond to isolated small forest patches. We reported the area of forest in each fragmentation category for the six years and analyzed the dynamics of fragmentation over the six decades. We also computed the distance to forest edge for all forest pixels for the years 1953, 1973, 1990, 2000, 2005, 2010 and 2014. For that, we used the function `gdal_proximity.py` of the GDAL software (http://www.gdal.org/). We computed the mean and 95% quantiles of the distance to forest edge and looked at the evolution of these values with time.

## 3. Results

### 3.1 Dynamics of forest cover and deforestation intensity

Natural forests in Madagascar covered 16.0 Mha in 1953, about 27% of the national territory of 587,041 km2. In 2014, the forest cover dropped to 8.9 Mha, corresponding to about 15% of the national territory (Fig. 2 and Tab. 1). Madagascar has lost 44% and 37% of its natural forests between 1953 and 2013, and between c. 1973 and 2014 respectively (Fig. 2 and Tab. 1). In 2014 the remaining 8.9 Mha of natural forest were distributed as: 4.4 Mha of moist forest (50% of total forest cover), 2.6 Mha of dry forest (29%), 1.7 Mha of spiny forest (19%) and 0.18 Mha (2%) of mangrove forest (Fig. 1 and Tab. 2). Regarding the deforestation trend, we observed a progressive decrease of the deforestation rate after1990 from 205,000 ha/yr (1.63%/yr) over the period 1973-1990 to 44,300 ha/yr (0.43%/yr) over the period 2000-2005 (Tab. 1). Then from 2005, the deforestation rate has progressively increased and has more than doubled over the period 2010-2014 (98,700 ha/yr, 1.08%/yr) compared to 2000-2005 (Tab. 1). The deforestation trend characterized by a progressive decrease of the deforestation rate over the period 1990-2005 and a progressive increase of the deforestation after 2005 is valid for all four ecoregions (Tab. 3), with the exception of the spiny forest domain for which the deforestation rate during the period 2010-2013 was lower than during 2005-2010 (Tab. 3).

### 3.2 Comparison with previous forest-cover change studies in Madagascar

Forest-cover maps provided by previous studies over Madagascar were not exhaustive (unclassified areas) due to the presence of clouds in satellite imagery used to produce such maps. In @Harper2007, the maps of years 1990 and 2000 include 0.5 and 1.12 Mha of unknown cover type respectively. Proportions of unclassified areas are not reported in the two other existing studies by @MEFT2009 and @ONE2015. With our approach, we produced wall to wall forest-cover change maps from 1990 to 2014 for the full territory of Madagascar (Tab. 1). This allowed us to produce more robust estimates of forest-cover and deforestation rates over this period. Our forest-cover estimates over the period 1953-2013 (considering forest cover estimates at national level and by ecoregions for the 6 available dates) were well correlated  (0.99) to estimates from the three previous studies (Tab. 2) with a RMSE of 300,000 ha (6% of the mean forest cover of 4.8 Mha when considering all dates and forest types together). These small differences can be partly attributed to differences in ecoregion boundaries. Despite significant differences in deforestation estimates (Tab. 3), a similar deforestation trend was observed across studies with a decrease of deforestation rates over the period 1990-2005, followed by a progressive increase of the deforestation after 2005.

### 3.3 Evolution of forest fragmentation with time

>>>>

---
title: "Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
  bookdown::pdf_document2:
    #includes:
    #  before_body: doc_prefix.tex
    #  in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    toc: no
bibliography: bib/biblio.bib
fontsize: 12pt
link-citations: yes
csl: bib/journal-of-applied-ecology.csl
---

@@

The island of Madagascar has a unique biodiversity, mainly located in the tropical forests of the island. This biodiversity is highly threatened by anthropogenic deforestation. Existing historical forest maps at national level are scattered and have substantial gaps which prevent an exhaustive assessment of long-term deforestation trends in Madagascar. In this study, we combine historical national forest cover maps (covering the period 1953-2000) with a recent global annual tree cover loss dataset (2001-2014) to look at six decades of deforestation and forest fragmentation in Madagascar (from 1953 to 2014). We produced new forest cover maps at 30 m resolution for the year 1990 and annually from 2000 to 2014 over the full territory of Madagascar. We estimated that Madagascar has lost 44% of its natural forest cover over the period 1953-2014 (including 37% over the period 1973-2014). Natural forests cover 8.9 Mha in 2014 (15% of the national territory) and include 4.4 Mha (50%) of moist forests, 2.6 Mha (29%) of dry forests, 1.7 Mha of spiny forests (19%) and 177,000 ha (2%) of mangroves. Since 2005, the annual deforestation rate has progressively increased in Madagascar to reach 99,000 ha/yr during 2010-2014 (corresponding to a rate of 1.1%/yr). This increase is probably due to rapid population growth (close to 3%/yr) and to poor law enforcement in the country. Around half of the forest (46%) is now located at less than 100 m from the forest edge. Accurate forest cover change maps can be used to assess the effectiveness of past and current conservation programs and implement new strategies for the future. In particular, forest maps and estimates can be used in the REDD+ framework which aims at "Reducing Emissions from Deforestation and Forest Degradation" and for optimizing the current protected area network.

**Keywords**: biodiversity, climate-change, deforestation, forest-fragmentation, Madagascar, tropical forest.

## Introduction

Separated from the African continent and the Indian plate about 165 and
88 million years ago respectively [@Ali2008], the flora and fauna of
Madagascar followed its own evolutionary path. Isolation combined with a
high number of micro-habitats [@Pearson2009] has led to Madagascar's
exceptional biodiversity both in term of number of species and endemism
in many taxonomic groups [@Crottini2012; @Goodman2005]. Most of the
biodiversity in Madagascar is concentrated in the tropical forests of
the island which can be divided into four types: the moist forest in the
East, the dry forest in the West, the spiny forest in the South and the
mangroves on the West coast [@Vieilledent2016]. This unparalleled
biodiversity is severely threatened by deforestation
[@Harper2007; @Vieilledent2013] associated with human activities such as
slash-and-burn agriculture and pasture [@Scales2011]. Tropical forests
in Madagascar also store a large amount of carbon [@Vieilledent2016] and
high rates of deforestation in Madagascar are responsible for large
CO$_2$ emissions in the atmosphere [@Achard2014]. Deforestation
threatens species survival by directly reducing their available habitat
[@Brooks2002; @Tidd2001]. Forest fragmentation can also lead to species
extinction by isolating populations from each other and creating forest
patches too small to maintain viable populations [@Saunders1991].
Fragmentation also increases forest edge where ecological conditions
(such as air temperature, light intensity and air moisture) can be
dramatically modified, with consequences on the abundance and
distribution of species [@Murcia1995]. Forest fragmentation can also
have substantial effects on forest carbon storage capacity, as carbon
stocks are much lower at the forest edge than under a closed canopy
[@Brinck2017]. Moreover, forest carbon stocks vary spatially due to
climate or soil factors [@Saatchi2011; @Vieilledent2016]. As a
consequence, accurate and spatially explicit maps of forest cover and
forest cover change are necessary to monitor biodiversity loss and
carbon emissions from deforestation and forest fragmentation, assess the
efficiency of present conservation strategies [@Eklund2016], and
implement new strategies for the future
[@Vieilledent2013; @Vieilledent2016]. Simple time-series of forest cover
estimates, such as those provided by the FAO Forest Resource Assessment
report [@Keenan2015] are not sufficient.

Unfortunately, accurate and exhaustive forest cover maps are not
available for Madagascar after year 2000. @Harper2007 produced maps of
forest cover and forest cover changes over Madagascar for the years
_c._ 1953, _c._ 1973, 1990 and 2000. The _c._ 1953 forest map was
derived from the visual interpretation of aerial photography at coarse
scale (1/1,000,000). Forest maps for the years _c._ 1973, 1990, and 2000
were obtained from supervised classification of Landsat satellite images
at 60 m resolution (for the year 1973) or 30 m resolution (for years
1990 and 2000) and can be used to derive more accurate estimates of
forest cover (89.5% accuracy reported for the forest/non-forest map of
year 2000). Nonetheless, maps provided by @Harper2007 are not exhaustive
(due to the presence of clouds in the satellite imagery), e.g. 11
244 km2 are mapped as unknown cover type for the year 2000. Using a
similar supervised classification approach as in @Harper2007, more
recent maps have been produced for the periods 2000-2005-2010 by
national institutions, with the technical support of international
environmental NGOs [@MEFT2009; @ONE2013]. Another set of recent forest
cover maps using an advanced statistical tool for classification, the
Random Forest classifier [@Grinand2013; @Rakotomalala2015], was produced
for the periods 2005-2010-2013 [@ONE2015]. However, these maps are
either too old to give recent estimates of deforestation
[@MEFT2009; @ONE2013], include large areas of missing information due to
images with high percentage of cloud cover [@ONE2013], or show large
mis-classification in specific areas, especially in the dry and spiny
forest domain for which the spectral answer has a strong seasonal
behavior due to the deciduousness of such forests (overall accuracy is
lower than 0.8 for the dry and spiny forests for the maps produced by
@ONE2015). Moreover, the production of such forest maps from a
supervised classification approach requires significant resources,
especially regarding the image selection step (required to minimize
cloud cover) and the training step (visual interpretation of a large
number of polygons needed to train the classification algorithm)
[@Rakotomalala2015]. Most of this work of image selection and visual
interpretation would need to be repeated to produce new forest maps in
the future using a similar approach.

Global forest or tree cover products have also been published recently
and can be tested at the national scale for Madagascar. @Kim2014
produced a global forest cover change map from 1990 to 2000 (derived
from Landsat imagery). This product was updated to cover the period
1975-2005 (<http://glcf.umd.edu/data/landsatFCC/>) but forest cover maps
after 2005 were not produced. Moreover, the approach used in @Kim2014
did not accurately map the forests in the dry and spiny ecosystems of
Madagascar (see Fig. 8 in @Kim2014). @Hansen2013 mapped tree cover
percentage, annual tree cover loss and gain from 2000 to 2012 at global
scale at 30 m resolution. This product has since been updated and is now
available up to the year 2014 [@Hansen2013]. To map forest cover from
the @Hansen2013 product, a tree cover threshold must be selected (that
defines forest cover). Selecting such a threshold is not straightforward
as the accuracy of the global tree cover map strongly varies between
forest types, and is substantially lower for dry forests than for moist
forests [@Bastin2017]. Moreover, the @Hansen2013 product does not
provide information on land-use. In particular the global tree cover map
does not separate tree plantations such as oil palm or eucalyptus
plantations from natural forests [@Tropek2014]. Thus, the global tree
cover map from @Hansen2013 cannot be used alone to produce a map of
forest cover [@Tyukavina2017].

In this study, we present a simple approach which combines the
historical forest maps from @Harper2007 and more recent global products
from @Hansen2013 to derive annual wall-to-wall forest cover change maps
over the period 2000-2014 for Madagascar. We use the forest cover map
provided by @Harper2007 for the year 2000 (defining the land-use) with
the tree cover loss product provided by @Hansen2013 that we apply only
inside forest areas identified by @Harper2007. Similar to the approach
of @Harper2007, we also assess trends in deforestation rates and forest
fragmentation from _c._ 1953 to 2014. We finally discuss the possibility
to extend our approach to other tropical countries or repeat it in the
future. We also discuss how our results could help assess the
effectiveness of current conservation strategies, and implementation new
conservation strategies for the future in Madagascar.

## Materials and Methods

### Creation of new forest cover maps of Madagascar from 1953 to 2014

We produced annual forest/non-forest maps at 30 m resolution for the
full territory of Madagascar for the period 2000-2014 by combining the
forest map of year 2000 from @Harper2007, and the tree cover percentage
and annual tree cover loss maps over the period 2000-2014 from
@Hansen2013. The 2000 Harper's forest map includes 208,000 ha of
unclassified areas due to the presence of clouds on satellite images,
mostly (88%) within the moist forest domain which covered 4.17 Mha in
total in 2000. To provide a label (forest or non-forest) to these
unclassified pixels, we used the 2000 tree cover percentage map of
@Hansen2013 by selecting a threshold of 75% tree cover to define forest
cover as recommended by other studies for the moist domain
[@Achard2014; @Aleman2017]. To do so, the Hansen's 2000 tree cover map
was resampled on the same grid as the original Harper's map at 30 m
resolution using a bilinear interpolation. We thus obtained a forest
cover map for the year 2000 covering the full territory of Madagascar.
We then combined this forest cover map of the year 2000 with the annual
tree cover loss maps from 2001 to 2014 provided by @Hansen2013 to create
annual forest cover maps from 2001 to 2014 at 30 m resolution. To do so,
Hansen's tree cover loss maps were resampled on the same grid as the
original Harper's map at 30 m resolution using a nearest-neighbor
interpolation. We also completed the Harper's forest map of year 1990 by
filling unclassified areas (due to the presence of clouds on satellite
images) using our forest cover map of year 2000. To do so, we assumed
that if forest was present in 2000, the pixel was also forested in 1990.
The remaining unclassified pixels were limited to a relatively small
total area of _c._ 8,000 ha. We labeled these residual pixels as
non-forest, as for the year 2000. Similarly we completed the Harper's
forest map of year 1973 by filling unclassified areas using our forest
cover map of the year 1990 assuming that if forest was present in 1990,
it was also present in 1973. Contrary to the year 1990, the remaining
unclassified pixels for year 1973 corresponded to a significant total
area of 3.3 million ha. We also reprojected the forest cover map of year
1953 to a common projection in order to compare the forest cover area in
1953 with forest cover areas at the following dates. This map was
produced by scanning a paper map derived from aerial photos, and thus
could not be perfectly aligned with the other maps produced through
digital processing of satellite imagery [@Harper2007]. Finally for all
forest cover maps from 1973, the isolated single non-forest pixels
(i.e. fully surrounded by forest pixels) were removed, assuming that
single non-forest pixels inside a forest patch were not corresponding to
deforestation (they might correspond to selective logging activities).
This allowed us to avoid counting very small scale events (\<0.1 ha,
such as selective logging) as forest fragmentation. All the resulting
maps have been made permanently and freely available on the Zenodo
research data repository at <https://doi.org/10.5281/zenodo.1145785>.

### Computing forest cover areas and deforestation rates

From these new forest cover maps, we calculated the total forest cover
area for seven available years (1953-1973-1990-2000-2005-2010-2014), and
the annual deforested area and annual deforestation rate for the
corresponding six time periods between 1953 and 2014. The annual
deforestation rates were calculated using Eq. \@ref(eq:theta) [@Puyravaud2003; @Vieilledent2013]:

\begin{equation}
  \theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}]
  (\#eq:theta)
\end{equation}

In Eq. \ref{eq:theta}, $\theta$ is the annual deforestation rate (in
%/yr), $F_{t_2}$ and $F_{t_1}$ are the forest cover free of clouds at
both dates $t_2$ and $t_1$, and $t_2-t_1$ is the time-interval (in
years) between the two dates.

Because of the large unclassified area (3.3 million ha) in 1973, the
annual deforestation areas and rates for the two periods 1953-1973 and
1973-1990 are only partial estimates computed on the basis of the
available forest extent. Area and rate estimates are produced at the
national scale and for the four forest ecosystems present in Madagascar:
moist forest in the East, dry forest in the West, spiny forest in the
South, and mangroves on the Western coast
(Fig. \ref{fig:ecoregions}). To define the forest domains, we used a map
from the MEFT (*"Ministère de l'Environnement et des Forêts à
Madagascar"*) with the boundaries of the four ecoregions in Madagascar.
Ecoregions were defined on the basis of climatic and vegetation criteria
using the climate classification by @Cornet1974 and the vegetation
classification from the 1996 IEFN national forest inventory [@IEFN1996].
Because mangrove forests are highly dynamic ecosystems that can expand
or contract on decadal scales depending on changes in environmental
factors [@Armitage2015], a fixed delimitation of the mangrove ecoregion
on six decades might not be fully appropriate. As a consequence, our
estimates of the forest cover and deforestation rates for mangroves in
Madagascar must be considered with this limitation.

### Comparing our forest cover and deforestation rate estimates with previous studies

We compared our estimates of forest cover and deforestation rates with
estimates from the three existing studies at the national scale for
Madagascar: (i) [@Harper2007], (ii) [@MEFT2009] and (iii) [@ONE2015].
@Harper2007 provides forest cover and deforestation estimates for the
periods c. 1953-c. 1973-1990-2000. @MEFT2009 provides estimates for the
periods 1990-2000-2005 and @ONE2015 provides estimates for the periods
2005-2010-2013. To compare our forest cover and deforestation estimates
over the same time periods, we consider an additional time-period in our
study (2010-2013) by creating an extra forest cover map for the year
2013. We computed the Pearson's correlation coefficient and the root
mean square error (RMSE) between our forest cover estimates and forest
cover estimates from previous studies for all the dates and forest types
(including also the total forest cover estimates). For previous studies,
the computation of annual deforestation rates (in %/yr) is not always
detailed and might slightly differ from one study to another [see @Puyravaud2003]. @Harper2007 also provide total deforested areas for the
two periods 1973-1990 and 1990-2000. We converted these values into
annual deforested area estimates. When annual deforested areas were not
reported (for 1953-1973 in @Harper2007 and in @MEFT2009 and @ONE2015),
we computed them from the forest cover estimates in each study. These
estimates cannot be corrected from the potential bias due to the
presence of residual clouds. forest cover and deforestation rates were
then compared between all studies for the whole of Madagascar and the
four ecoregions. The same ecoregion boundaries as in our study were used
in @ONE2015 but this was not the case for @Harper2007 and @MEFT2009,
which can explain a part of the differences between the estimates.

### Fragmentation

We also conducted an analysis of changes in forest fragmentation for the
years 1953, 1973, 1990, 2000, 2005, 2010 and 2014 at 30 m resolution. We
used a moving window of $51 \times 51$ pixels centered on each forest
pixel to compute the percentage of forest pixels in the neighborhood. We
used this percentage as an indication of the forest fragmentation. The
size of the moving windows was based on a compromise: a sufficiently
high number of cells (here 2601) had to be considered to be able to
compute a percentage and a reasonably low number of cells had to be
choosen to have a local estimate of the fragmentation. Computations were
done using the function `r.neighbors` of the GRASS GIS software
[@Neteler2008]. Using the density of forest in the neighborhood, we
defined five forest fragmentation classes: 0-20% (highly fragmented),
21-40%, 41-60%, 61-80% and 81-100% (lowly fragmented). We reported the
percentage of forest falling in each fragmentation class for the six
years and analyzed the dynamics of fragmentation over the six decades.

We also computed the distance to forest edge for all forest pixels for
the years 1953, 1973, 1990, 2000, 2005, 2010 and 2014. For that, we used
the function `gdal_proximity.py` of the GDAL library
(<http://www.gdal.org/>). We computed the mean and 90% quantiles (5% and
95%) of the distance to forest edge and looked at the evolution of these
values with time. Previous studies have shown that forest micro-habitats
were mainly altered within the first 100 m of the forest edge
[@Brinck2017; @Gibson2013; @Murcia1995; @Broadbent2008]. Consequently,
we also estimated the percentage of forest within the first 100 m of the
forest edge for each year and looked at the evolution of this percentage
over the six decades.

## Results

### Forest cover change and deforestation rates

Natural forests in Madagascar covered 16.0 Mha in 1953, about 27% of
the national territory of 587,041 km2. In 2014, the forest cover
dropped to 8.9 Mha, corresponding to about 15% of the national
territory (Fig.~\ref{fig:fcc} and
Tab. \@ref(tab:forest_cover)). Madagascar has lost 44% and 37% of its
natural forests between 1953 and 2014, and between 1973 and 2014
respectively (Fig.~\ref{fig:fcc} and Tab. \@ref(tab:forest_cover)). In
2014 the remaining 8.9 Mha of natural forest were distributed as
follow: 4.4 Mha of moist forest (50% of total forest cover), 2.6 Mha
of dry forest (29%), 1.7 Mha of spiny forest (19%) and 0.18 Mha
(2%) of mangrove forest (Fig.~\ref{fig:ecoregion} and
Tab. \@ref(tab:comp_forest)). Regarding the deforestation trend, we
observed a progressive decrease of the deforestation rate after 1990
from 205,000 ha/yr (1.6%/yr) over the period 1973-1990 to 42,000
ha/yr (0.4%/yr) over the period 2000-2005
(Tab. \@ref(tab:forest_cover)). Then from 2005, the deforestation rate
has progressively increased and has more than doubled over the period
2010-2014 (99,000 ha/yr, 1.1%/yr) compared to 2000-2005
(Tab. \@ref(tab:forest_cover)). The deforestation trend, characterized
by a progressive decrease of the deforestation rate over the period
1990-2005 and a progressive increase of the deforestation after 2005,
is valid for all four forest types except the spiny forest
(Tab. \@ref(tab:comp_defor)). For the spiny forest, the deforestation
rate during the period 2010-2013 was lower than on the period
2005-2010 (Tab. \@ref(tab:comp_defor)).

### Comparison with previous forest cover change studies in Madagascar
